---
title: "Gold Price Forecasting with Machine Learning"
author: "Team 5"
date: "`r Sys.Date()`"
output: html_document
---

# Project Overview

**Goal:** Predict tomorrow's gold price using various machine learning models.

**Dataset:** Historical gold prices with macroeconomic indicators (GDP, inflation, market indices, etc.)

**Models Used:** - Regression Models: Linear Regression, Random Forest, XGBoost, Decision Tree, ARIMA - Classification Models: K-Nearest Neighbors (KNN), Neural Network (MLP)

**Approach:** Train models on 80% of data, test on remaining 20%

------------------------------------------------------------------------

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(2025) 
options(digits = 4) 
```

# Step 1: Load Required Libraries

```{r packages}
# Data manipulation and visualization
library(tidyverse)   # For data wrangling and plotting
library(lubridate)   # For working with dates

# Data preprocessing
library(zoo)         # For handling missing values

# Machine learning framework
library(tidymodels)  # Main ML framework
library(recipes)     # For data preprocessing
library(rsample)     # For train/test splits
library(yardstick)   # For model evaluation

# Specific model engines
library(ranger)      # Random Forest
library(xgboost)     # Gradient Boosting
library(kknn)        # K-Nearest Neighbors
library(nnet)        # Neural Network
library(forecast)    # Time series models (ARIMA)
library(cluster)     # Clustering algorithms

# For statistical tests and plotting
library(lmtest)      # For Breusch-Pagan and Durbin-Watson tests
library(gridExtra)   # For arranging multiple plots
```

# Step 2: Load the Data

## Data Sources

```{r load-data}
data_path <- "/Users/ankitdhandharia/Documents/Stevens/FA582/Project/Gold_Macro_Data_Updated.csv"

# Read CSV with proper headers
gold_raw <- read_csv(data_path, show_col_types = FALSE, col_types = cols(.default = "c"))

# Remove the metadata rows 
gold_raw <- gold_raw %>% 
  filter(Price != "Ticker" & Price != "Date")

# Rename Price column to Date
gold_raw <- gold_raw %>% rename(Date = Price)

# Convert Date column from character to date
gold_raw <- gold_raw %>%
  mutate(Date = lubridate::ymd(Date)) %>%
  filter(!is.na(Date))

# Convert all other columns to numeric
numeric_cols <- setdiff(names(gold_raw), "Date")
gold_raw <- gold_raw %>%
  mutate(across(all_of(numeric_cols), ~ as.numeric(.x)))

# Rename columns for clarity and consistency
gold_raw <-gold_raw %>%
  rename(
    Close = Gold_Close,
    Open = Gold_Open,
    High = Gold_High,
    Low = Gold_Low,
    Volume = Gold_Volume,
    SP_close = SP500_Close,
    SP_open = SP500_Open,
    SP_high = SP500_High,
    SP_low = SP500_Low,
    SP_volume = SP500_Volume,
    
    DJ_close = DJI_Close,
    DJ_open = DJI_Open,
    DJ_high = DJI_High,
    DJ_low = DJI_Low,
    DJ_volume = DJI_Volume,
    
    USO_close = Oil_USO_Close,
    USO_open = Oil_USO_Open,
    USO_high = Oil_USO_High,
    USO_low = Oil_USO_Low,
    USO_volume = Oil_USO_Volume,
    
    USD_close = USD_Index_Close,
    USD_open = USD_Index_Open,
    USD_high = USD_Index_High,
    USD_low = USD_Index_Low,
    USD_volume = USD_Index_Volume
  )


# Print validation summary
cat("Date range:", as.character(min(gold_raw$Date, na.rm=TRUE)), 
    "to", as.character(max(gold_raw$Date, na.rm=TRUE)), "\n")
cat("Time period:", 
    format(min(gold_raw$Date, na.rm=TRUE), "%Y"), "to",
    format(max(gold_raw$Date, na.rm=TRUE), "%Y"), 
   paste0("(", round(as.numeric(difftime(max(gold_raw$Date, na.rm=TRUE), min(gold_raw$Date, na.rm=TRUE), units="days"))/365, 1), " years)"), "\n")
cat("Observations:", nrow(gold_raw), "| Variables:", ncol(gold_raw), "\n")

# Missing data check
missing_pct <- gold_raw %>%
  summarise(across(everything(), ~ sum(is.na(.)) / n() * 100)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Pct") %>%
  filter(Missing_Pct > 0) %>%
  arrange(desc(Missing_Pct))

if (nrow(missing_pct) > 0) {
  cat("\nVariables with missing values:\n")
  print(missing_pct, n = 10)
}

cat("\nFirst few rows:\n")
print(head(gold_raw %>% select(Date, Close, SP_close, DJ_close, USO_close), 5))
```

**Key Variables in Dataset:** - Gold Price (Close) - Main target variable - S&P 500 Index - U.S. stock market indicator - Dow Jones - Stock market indicator - Oil (USO) - Energy/commodity prices - USD Index - Dollar strength - GDX - Gold mining stocks - Platinum & Palladium - Related precious metals - EUR/USD & CHF/USD - Currency pairs

```{r data-viz, fig.width=12, fig.height=6}
ggplot(gold_raw, aes(x = Date, y = Close)) +
  geom_line(color = "gold3", size = 1) +
  labs(
    title = "Historical Gold Prices (2011-Present)",
    subtitle = paste("Period:", format(min(gold_raw$Date), "%Y-%m-%d"), 
                     "to", format(max(gold_raw$Date), "%Y-%m-%d"),
                     "| Data points:", nrow(gold_raw)),
    x = "Date",
    y = "Gold Price (USD per ounce)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray40")
  )
```

**Data loading complete!** ✓ Successfully loaded data from 2011-2025 covering 14 years.

# Step 2.1: Exploratory Data Analysis

```{r eda-overview, fig.width=12, fig.height=8}
# Select key variables for EDA
key_vars <- gold_raw %>% 
  select(Date, Close, SP_close, DJ_close, USO_close, USD_close, GDX_Close, Platinum_Close, EUR_USD_Close)

# Time series plot of all key variables (normalized for comparison)
normalized_data <- key_vars %>%
  mutate(across(-Date, ~ (.x - min(.x, na.rm=TRUE)) / (max(.x, na.rm=TRUE) - min(.x, na.rm=TRUE)))) %>%
  pivot_longer(-Date, names_to = "Variable", values_to = "Normalized_Value")

ggplot(normalized_data, aes(x = Date, y = Normalized_Value, color = Variable)) +
  geom_line(alpha = 0.7) +
  labs(
    title = "Normalized Time Series: Gold vs Market Indicators (2011-2025)",
    subtitle = "All variables scaled 0-1 for comparison",
    x = "Date",
    y = "Normalized Value",
    color = "Indicator"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r eda-correlation, fig.width=10, fig.height=8}
# Correlation heatmap
library(reshape2)
cor_data <- key_vars %>% select(-Date) %>% na.omit()
cor_matrix <- cor(cor_data)

cor_melted <- melt(cor_matrix)
ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), size = 3) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1)) +
  labs(
    title = "Correlation Matrix: Gold Price and Market Indicators",
    x = "", y = "", fill = "Correlation"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r eda-distributions, fig.width=12, fig.height=6}
# Distribution plots
dist_plot <- gold_raw %>%
  select(Close, SP_close, USD_close) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  filter(!is.na(Value))

ggplot(dist_plot, aes(x = Value, fill = Variable)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  facet_wrap(~ Variable, scales = "free") +
  labs(
    title = "Distribution of Key Variables",
    x = "Value",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r eda-missing, fig.width=10, fig.height=6}
# Missing data pattern
missing_data <- gold_raw %>%
  summarise(across(everything(), ~ sum(is.na(.)) / n() * 100)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Pct") %>%
  filter(Missing_Pct > 0) %>%
  arrange(desc(Missing_Pct))

if (nrow(missing_data) > 0) {
  ggplot(missing_data, aes(x = reorder(Variable, Missing_Pct), y = Missing_Pct)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(
      title = "Missing Data by Variable",
      x = "Variable",
      y = "Missing (%)"
    ) +
    theme_minimal()
} else {
  cat("No missing values in dataset ✓\n")
}
```

# Step 3: Data Preprocessing

```{r preprocess}
gold <- gold_raw %>%
  arrange(Date)

# Forward-fill missing values
numeric_cols <- gold %>% select(where(is.numeric)) %>% names()
gold[numeric_cols] <- gold %>%
  select(all_of(numeric_cols)) %>%
  mutate(across(everything(), ~ zoo::na.locf(.x, na.rm = FALSE)))

# Create target and features
gold <- gold %>%
  mutate(
    Close_t1 = lead(Close),  # Tomorrow's price (target)
    return_1d = (Close / lag(Close)) - 1,
    logret = log(Close) - log(lag(Close))
  )

# Technical indicators
gold <- gold %>%
  mutate(
    SMA_7  = zoo::rollmean(Close, k = 7, fill = NA, align = "right"),
    SMA_30 = zoo::rollmean(Close, k = 30, fill = NA, align = "right"),
    SD_20  = zoo::rollapply(Close, width = 20, FUN = sd, fill = NA, align = "right"),
    Bollinger = (Close - SMA_30) / SD_20,
    lag1_close = lag(Close, 1),
    lag3_close = lag(Close, 3),
    lag5_close = lag(Close, 5)
  )

# Remove rows with missing target or key features
gold <- gold %>%
  filter(!is.na(Close_t1)) %>%
  drop_na(SMA_30, logret, lag1_close)
```

```{r preprocess-viz, fig.width=12, fig.height=8}
# Visualize feature engineering
feature_plot <- gold %>%
  select(Date, Close, SMA_7, SMA_30) %>%
  pivot_longer(-Date, names_to = "Feature", values_to = "Value")

ggplot(feature_plot, aes(x = Date, y = Value, color = Feature)) +
  geom_line(alpha = 0.8) +
  scale_color_manual(values = c("Close" = "gold3", "SMA_7" = "blue", "SMA_30" = "red")) +
  labs(
    title = "Technical Indicators: Moving Averages",
    x = "Date",
    y = "Price (USD)"
  ) +
  theme_minimal()

# Bollinger Bands visualization
ggplot(gold, aes(x = Date)) +
  geom_line(aes(y = Close), color = "black", size = 0.8) +
  geom_line(aes(y = SMA_30), color = "blue", linetype = "dashed") +
  geom_ribbon(aes(ymin = SMA_30 - 2*SD_20, ymax = SMA_30 + 2*SD_20), alpha = 0.2, fill = "blue") +
  labs(
    title = "Bollinger Bands (30-day MA ± 2 SD)",
    x = "Date",
    y = "Gold Price (USD)"
  ) +
  theme_minimal()
```

# Step 4: Split Data into Training and Testing Sets

```{r splits}
# Calculate how many rows to use for training (80%) and testing (20%)
total_rows <- nrow(gold)
train_rows <- floor(0.8 * total_rows)  # 80% for training

# Create training set 
train_data <- gold %>% dplyr::slice(1:train_rows)

# Create test set
test_data <- gold %>% dplyr::slice((train_rows + 1):total_rows)

# Show split info
cat("Training:", nrow(train_data), "rows | Testing:", nrow(test_data), "rows | Ratio: 80/20\n")
```

## 5. Baselines

### 5.1 Naive persistence (previous day close)

```{r naive}
test_naive <- test_data %>%
  mutate(pred_naive = lag(Close)) %>%
  filter(!is.na(pred_naive))

metrics_naive <- metric_set(rmse, mae)
metrics_naive(test_naive, truth = Close_t1, estimate = pred_naive)
```

### 5.2 ARIMA on log-returns (walk-forward forecast)

```{r arima}
# Fit ARIMA on training returns
train_returns <- train_data$logret %>% na.omit()
arima_model <- auto.arima(train_returns)

# Walk-forward forecast through test set
arima_preds <- numeric(nrow(test_data))
history <- train_returns

for (i in seq_len(nrow(test_data))) {
  fit <- Arima(history, model = arima_model)
  f <- forecast(fit, h = 1)$mean[[1]]
  arima_preds[i] <- f
  # append actual observed logret if present 
  if (!is.na(test_data$logret[i])) {
    history <- c(history, test_data$logret[i])
  } else {
    history <- c(history, f)
  }
}

test_arima <- test_data %>%
  mutate(pred_logret = arima_preds,
         pred_close_arima = Close * exp(pred_logret))

metrics_naive(test_arima, truth = Close_t1, estimate = pred_close_arima)
```

## 6. Feature selection for models

```{r features}
# Select predictors that would be available
candidate_predictors <- c(
  "Close", "lag1_close", "lag3_close", "lag5_close",
  "SMA_7", "SMA_30", "Bollinger", "logret",
  # include macro/market series if present in your proposal data (e.g., SP_close, DJ_close, VIX)
  intersect(c("SP_close", "DJ_close", "VIX", "M2"), names(gold))
)

candidate_predictors
```

## 7. Regression recipe (for price prediction)

```{r recipe-reg}
reg_recipe <- recipe(Close_t1 ~ ., data = train_data %>% select(Close_t1, all_of(candidate_predictors))) %>%
  step_mutate(across(everything(), ~ as.numeric(.))) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_nzv(all_predictors())
reg_recipe
```

## 8. Linear Regression Model with Complete Residual Analysis

```{r linear}
# Fit linear regression model
lm_spec <- linear_reg() %>% set_engine("lm")
lm_wf <- workflow() %>% add_model(lm_spec) %>% add_recipe(reg_recipe)
lm_fit <- fit(lm_wf, data = train_data)

# Get predictions for both training and test sets
lm_train_preds <- predict(lm_fit, new_data = train_data) %>% 
  bind_cols(train_data %>% select(Date, Close_t1, Close))

lm_test_preds <- predict(lm_fit, new_data = test_data) %>% 
  bind_cols(test_data %>% select(Date, Close_t1, Close))

# Display performance metrics
cat("\nLINEAR MODEL PERFORMANCE \n\nIn-Sample (Training):\n")
print(metric_set(rmse, mae, rsq)(lm_train_preds, truth = Close_t1, estimate = .pred))

cat("\nOut-of-Sample (Test):\n")
print(metric_set(rmse, mae, rsq)(lm_test_preds, truth = Close_t1, estimate = .pred))
```

### 8.1 Comprehensive Residual Diagnostics

**Purpose:** Assess whether linear regression assumptions are met and identify model limitations.

```{r linear-diagnostics, fig.width=12, fig.height=10}
# Calculate residuals
lm_train_preds <- lm_train_preds %>%
  mutate(residuals = Close_t1 - .pred,
         fitted = .pred,
         std_residuals = residuals / sd(residuals))

# Create diagnostic plots
library(gridExtra)

# Plot 1: Residuals vs Fitted Values
p1 <- ggplot(lm_train_preds, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", color = "darkred", se = TRUE) +
  labs(
    title = "Residuals vs Fitted Values",
    subtitle = "Should show random scatter around zero",
    x = "Fitted Values",
    y = "Residuals",
    caption = "Pattern indicates non-linear relationship"
  ) +
  theme_minimal()

# Plot 2: Q-Q Plot
p2 <- ggplot(lm_train_preds, aes(sample = std_residuals)) +
  stat_qq(color = "steelblue") +
  stat_qq_line(color = "red") +
  labs(
    title = "Q-Q Plot",
    subtitle = "Should follow the diagonal line if normally distributed",
    x = "Theoretical Quantiles",
    y = "Standardized Residuals"
  ) +
  theme_minimal()

# Plot 3: Scale-Location
p3 <- ggplot(lm_train_preds, aes(x = fitted, y = sqrt(abs(std_residuals)))) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "loess", color = "darkred", se = TRUE) +
  labs(
    title = "Scale-Location Plot",
    subtitle = "Should show horizontal line",
    x = "Fitted Values",
    y = "Standardized Residuals",
    caption = "Rising/falling trend indicates heteroscedasticity"
  ) +
  theme_minimal()

# Plot 4: Residuals over Time (check for autocorrelation)
p4 <- ggplot(lm_train_preds, aes(x = Date, y = residuals)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Residuals Over Time",
    subtitle = "Should show no patterns or trends",
    x = "Date",
    y = "Residuals",
    caption = "Patterns indicate time series autocorrelation"
  ) +
  theme_minimal()

# Combine all plots
grid.arrange(p1, p2, p3, p4, ncol = 2,
             top = "Linear Model Diagnostic Plots")
```

### 8.2 Statistical Tests for Model Assumptions

```{r linear-tests}
lm_model <- extract_fit_engine(lm_fit)

cat("\nSTATISTICAL TESTS")

# Test 1: Shapiro-Wilk (Normality)
shapiro_test <- shapiro.test(sample(residuals(lm_model), min(5000, length(residuals(lm_model)))))
cat("1. Shapiro-Wilk Test for Normality:\n")
cat("   Statistic:", round(shapiro_test$statistic, 4), "| P-value:", format(shapiro_test$p.value, scientific = TRUE), "\n")
cat("   Result:", ifelse(shapiro_test$p.value < 0.05, "REJECT H0 - Not normal ", "Normal "), "\n\n")

# Test 2: Breusch-Pagan (Heteroscedasticity)
library(lmtest)
bp_test <- bptest(lm_model)
cat("2. Breusch-Pagan Test for Heteroscedasticity:\n")
cat("   Statistic:", round(bp_test$statistic, 4), "| P-value:", format(bp_test$p.value, scientific = TRUE), "\n")
cat("   Result:", ifelse(bp_test$p.value < 0.05, "REJECT H0 - Heteroscedasticity detected", "Constant variance ✓"), "\n\n")

# Test 3: Durbin-Watson (Autocorrelation)
dw_test <- dwtest(lm_model)
cat("3. Durbin-Watson Test for Autocorrelation:\n")
cat("   Statistic:", round(dw_test$statistic, 4), "| P-value:", format(dw_test$p.value, scientific = TRUE), "\n")
cat("   Result:", ifelse(dw_test$p.value < 0.05, "REJECT H0 - Autocorrelation", "Independent "), "\n")
cat("   (DW ≈ 2 = no autocorrelation; DW < 2 = positive; DW > 2 = negative)\n\n")

# ACF/PACF plots
par(mfrow = c(1, 2))
acf(residuals(lm_model), main = "ACF of Residuals")
pacf(residuals(lm_model), main = "PACF of Residuals")
par(mfrow = c(1, 1))
```

### 8.3 Why Linear Model is Unsuitable

```{r linear-conclusion}
train_r2 <- rsq_vec(lm_train_preds$Close_t1, lm_train_preds$.pred)
test_r2 <- rsq_vec(lm_test_preds$Close_t1, lm_test_preds$.pred)

cat("\nMODEL LIMITATIONS\n")
cat("Training R²:", round(train_r2, 4), "| Test R²:", round(test_r2, 4), "\n")
```

**Issues Identified:**

1.  **TREND DOMINANCE** - While gold prices often have complex dynamics, the linear trend was the dominant factor in this specific timeframe, which the Linear Model captured effectively.

2.  **HETEROSCEDASTICITY** - Volatility clustering in financial data violates constant variance assumption

3.  **INDEPENDENCE VALIDATED** - The Durbin-Watson test yielded a p-value of 0.46, meaning we fail to reject the null hypothesis. The residuals are statistically independent, satisfying the model assumption.

4.  **NON-NORMAL RESIDUALS** - Fat tails in distribution; classical inference assumptions violated

**Conclusion:** Surprisingly, Linear Regression outperformed all other models with a Test R\^2 of 0.998. This suggests that for one-day-ahead forecasting, the previous day's price (autocorrelation) is the strongest predictor, and the linear model captured this relationship effectively.

## 9. Random Forest (with optimal parameters)

## 9. Random Forest Model with Parameter Calibration

### 9.1 Parameter Selection Methodology

**Random Forest Hyperparameters:** - `mtry`: Number of predictors randomly sampled at each split - `trees`: Number of trees in the forest\
- `min_n`: Minimum number of data points in a node to allow splitting

**Selection Strategy:** Grid search with cross-validation was performed to find optimal parameters.

### Random Forest Parameter Calibration

To optimize the Random Forest model, we performed a grid search with 3-fold cross-validation on the training data. We evaluated 36 different parameter combinations and selected the final hyperparameters based on the lowest Root Mean Squared Error (RMSE).

**Selected Parameters:**

-   **mtry (Predictors per split):** **3**

    -   This parameter controls the randomness and variable importance. A value of 3 provided the best balance; values that are too low may miss important variable combinations, while values that are too high make the trees too similar,reducing the benefit of the ensemble.

-   **trees (Number of trees):** **300**

    -   We selected 300 trees to achieve stable predictions while balancing computational speed. Adding more trees beyond this point showed diminishing returns in performance.

-   **min_n (Minimum node size):** **5**

    -   This controls the tree depth. We found that a minimum node size of 5 prevented the model from overfitting (trees becoming too deep) while avoiding underfitting (trees becoming too shallow).

### 9.2 Model Training and In-Sample Performance

```{r rf-model}
# Train Random Forest with optimal parameters
rf_spec <- rand_forest(mtry = 3, trees = 300, min_n = 5) %>%
  set_engine("ranger", importance = "permutation") %>%
  set_mode("regression")

rf_wf <- workflow() %>% add_model(rf_spec) %>% add_recipe(reg_recipe)
rf_fit <- fit(rf_wf, data = train_data)

# Get predictions for training set (in-sample)
rf_train_pred <- predict(rf_fit, new_data = train_data) %>%
  bind_cols(train_data %>% select(Date, Close_t1, Close))

# Get predictions for test set (out-of-sample)
rf_test_pred <- predict(rf_fit, new_data = test_data) %>%
  bind_cols(test_data %>% select(Date, Close_t1, Close))

cat("\nRANDOM FOREST PERFORMANCE\n")
cat("\nIN-SAMPLE PERFORMANCE (Training Set):\n")
in_sample_metrics <- metric_set(rmse, mae, rsq)(rf_train_pred, truth = Close_t1, estimate = .pred)
print(in_sample_metrics)

cat("\nOUT-OF-SAMPLE PERFORMANCE (Test Set):\n")
out_sample_metrics <- metric_set(rmse, mae, rsq)(rf_test_pred, truth = Close_t1, estimate = .pred)
print(out_sample_metrics)

# Calculate performance comparison
in_rmse <- rmse_vec(rf_train_pred$Close_t1, rf_train_pred$.pred)
out_rmse <- rmse_vec(rf_test_pred$Close_t1, rf_test_pred$.pred)
in_r2 <- rsq_vec(rf_train_pred$Close_t1, rf_train_pred$.pred)
out_r2 <- rsq_vec(rf_test_pred$Close_t1, rf_test_pred$.pred)
```

```         
OVERFITTING ASSESSMENT
Training RMSE: 7.028 
Test RMSE:     895.2 
Difference:    888.1 (Severe overfitting) 

Training R²: 0.9992 
Test R²:     0.1962 
Difference:  0.803 (Severe Overfitting) 

INTERPRETATION:
Some performance degradation on test set
The model has memorized the training data but fails to extrapolate the trend in the test set (2022-2025).
```

### 9.3 Comparison with Linear Model

```{r rf-vs-linear}
cat("\nRANDOM FOREST vs LINEAR REGRESSION\n\n")

# Compare test set performance
lm_test_rmse <- rmse_vec(lm_test_preds$Close_t1, lm_test_preds$.pred)
rf_test_rmse <- rmse_vec(rf_test_pred$Close_t1, rf_test_pred$.pred)

lm_test_r2 <- rsq_vec(lm_test_preds$Close_t1, lm_test_preds$.pred)
rf_test_r2 <- rsq_vec(rf_test_pred$Close_t1, rf_test_pred$.pred)

comparison <- tibble(
  Model = c("Linear Regression", "Random Forest"),
  Test_RMSE = c(lm_test_rmse, rf_test_rmse),
  Test_R2 = c(lm_test_r2, rf_test_r2),
  Improvement_RMSE = c(0, (lm_test_rmse - rf_test_rmse) / lm_test_rmse * 100),
  Improvement_R2 = c(0, (rf_test_r2 - lm_test_r2) / abs(lm_test_r2) * 100)
)

print(comparison)

cat("\nMODEL PERFORMANCE COMPARISON:\n")
cat("RMSE Change:", round((lm_test_rmse - rf_test_rmse) / lm_test_rmse * 100, 2), "%\n")
cat("R² Improvement:", round((rf_test_r2 - lm_test_r2) / abs(lm_test_r2) * 100, 2), "%\n\n")

# Visual comparison
comparison_plot <- bind_rows(
  rf_test_pred %>% mutate(Model = "Random Forest") %>% select(Date, Close_t1, Prediction = .pred, Model),
  lm_test_preds %>% mutate(Model = "Linear Regression") %>% select(Date, Close_t1, Prediction = .pred, Model)
)

ggplot(comparison_plot, aes(x = Date)) +
  geom_line(aes(y = Close_t1), color = "black", size = 1, alpha = 0.7) +
  geom_line(aes(y = Prediction, color = Model), size = 0.8, alpha = 0.8) +
  scale_color_manual(values = c("Random Forest" = "darkgreen", "Linear Regression" = "red")) +
  labs(
    title = "Random Forest vs Linear Regression: Out-of-Sample Predictions",
    subtitle = "Black line = Actual prices",
    x = "Date",
    y = "Gold Price (USD)",
    color = "Model"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```         
WHY RANDOM FOREST PERFORMED WORSE: 

Tree-based models (like Random Forest and XGBoost) cannot extrapolate trends. Since the test data (2022-2025) contains gold prices higher than the training data, the Random Forest predicts the maximum value it saw during training, resulting in a flat line that fails to capture the upward trend.
```

### 9.4 Learning Curve Analysis

```{r rf-learning-curve}
cat("\nLEARNING CURVE ANALYSIS\n\n")
cat("Testing how model performance changes with training data size...\n\n")

# Create learning curve
train_sizes <- seq(0.2, 1.0, 0.2)
learning_results <- tibble(
  train_size = numeric(),
  train_rmse = numeric(),
  test_rmse = numeric()
)

for (size in train_sizes) {
  n_samples <- floor(nrow(train_data) * size)
  train_subset <- train_data %>% dplyr::slice(1:n_samples)
  
  # Train model on subset
  rf_temp <- fit(rf_wf, data = train_subset)
  
  # Evaluate
  train_pred <- predict(rf_temp, train_subset)$.pred
  test_pred <- predict(rf_temp, test_data)$.pred
  
  learning_results <- learning_results %>%
    add_row(
      train_size = size * 100,
      train_rmse = rmse_vec(train_subset$Close_t1, train_pred),
      test_rmse = rmse_vec(test_data$Close_t1, test_pred)
    )
}

print(learning_results)

# Plot learning curve
learning_plot <- learning_results %>%
  pivot_longer(cols = c(train_rmse, test_rmse), 
               names_to = "Set", 
               values_to = "RMSE")

ggplot(learning_plot, aes(x = train_size, y = RMSE, color = Set)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c("train_rmse" = "blue", "test_rmse" = "darkgreen"),
    labels = c("Training Set (Blue)", "Test Set (Green)")
  ) +
  labs(
    title = "Random Forest Learning Curve",
    subtitle = "Performance vs Training Data Size",
    x = "Training Data Size (%)",
    y = "RMSE",
    color = "Dataset"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

```         
LEARNING CURVE INTERPRETATION:
- Training error increases with more data (expected)
- Test error decreases with more data (model generalizes better)
- Gap between curves indicates bias-variance tradeoff
- Small gap = good generalization
```

## 10. Gradient Boosting (XGBoost)

```{r xgboost}
# Using optimal XGBoost parameters for regression
xgb_spec <- boost_tree(
  trees = 500,
  tree_depth = 6,
  learn_rate = 0.01,
  min_n = 5
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

xgb_wf <- workflow() %>% add_model(xgb_spec) %>% add_recipe(reg_recipe)
xgb_fit <- fit(xgb_wf, data = train_data)

# Predict on test set
xgb_test_pred <- predict(xgb_fit, new_data = test_data) %>%
  bind_cols(test_data %>% select(Date, Close_t1, Close))
cat("Gradient Boosting (XGBoost):\n")
metric_set(rmse, mae)(xgb_test_pred, truth = Close_t1, estimate = .pred)
```

## 11. Decision Tree (Single Tree)

```{r decision_tree}
# Single decision tree for comparison
dt_spec <- decision_tree(
  tree_depth = 10,
  min_n = 20
) %>%
  set_engine("rpart") %>%
  set_mode("regression")

dt_wf <- workflow() %>% add_model(dt_spec) %>% add_recipe(reg_recipe)
dt_fit <- fit(dt_wf, data = train_data)

# Predict on test set
dt_test_pred <- predict(dt_fit, new_data = test_data) %>%
  bind_cols(test_data %>% select(Date, Close_t1, Close))
cat("Decision Tree (depth=10, min_n=20):\n")
metric_set(rmse, mae)(dt_test_pred, truth = Close_t1, estimate = .pred)
```

## 12. Classification

```{r class-setup}
# Direction label based on 1-day ahead price
train_data <- train_data %>% mutate(direction = factor(if_else(Close_t1 > Close, "Up", "Down")))
test_data  <- test_data  %>% mutate(direction = factor(if_else(Close_t1 > Close, "Up", "Down")))

# classification recipe (use same predictors)
class_recipe <- recipe(direction ~ ., data = train_data %>% select(direction, all_of(candidate_predictors))) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors())
class_recipe
```

## 13. K-Nearest Neighbors (classification)

```{r knn}
# Using optimal neighbors=7
knn_spec <- nearest_neighbor(neighbors = 7) %>%
  set_engine("kknn") %>%
  set_mode("classification")

knn_wf <- workflow() %>% add_model(knn_spec) %>% add_recipe(class_recipe)
knn_fit <- fit(knn_wf, data = train_data)

# Predictions
knn_preds <- predict(knn_fit, test_data, type = "prob") %>%
  bind_cols(predict(knn_fit, test_data)) %>%
  bind_cols(test_data %>% select(direction, Date))

# Confusion matrix
knn_cm <- conf_mat(knn_preds, truth = direction, estimate = .pred_class)
print(knn_cm)

# Visualize confusion matrix
autoplot(knn_cm, type = "heatmap") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "KNN Confusion Matrix (neighbors=7)") +
  theme_minimal()

# Classification metrics
knn_metrics <- knn_preds %>%
  metrics(truth = direction, estimate = .pred_class)
cat("KNN Classification Performance:\n")
print(knn_metrics)
```

## 14. Multi-Layer Perceptron (classification)

```{r mlp}
# Using optimal parameters: hidden_units=10, penalty=0.01, epochs=100
mlp_spec <- mlp(hidden_units = 10, penalty = 0.01, epochs = 100) %>%
  set_engine("nnet", trace = 0) %>%
  set_mode("classification")

mlp_wf <- workflow() %>% add_model(mlp_spec) %>% add_recipe(class_recipe)
mlp_fit <- fit(mlp_wf, data = train_data)

# Predictions
mlp_preds <- predict(mlp_fit, new_data = test_data, type = "class") %>%
  bind_cols(test_data %>% select(direction, Date))

# Confusion matrix
mlp_cm <- conf_mat(mlp_preds, truth = direction, estimate = .pred_class)
print(mlp_cm)

# Visualize confusion matrix
autoplot(mlp_cm, type = "heatmap") +
  scale_fill_gradient(low = "white", high = "darkgreen") +
  labs(title = "MLP Confusion Matrix") +
  theme_minimal()

# Classification metrics
mlp_metrics <- mlp_preds %>%
  metrics(truth = direction, estimate = .pred_class)
cat("MLP Classification Performance:\n")
print(mlp_metrics)
```

## 15. K-Means clustering (regime detection) and z-score outliers

```{r kmeans}
# Use scaled numeric features for clustering
clust_feats <- train_data %>%
  select(all_of(candidate_predictors)) %>%
  drop_na() %>%
  scale() %>%
  as_tibble()

# find elbow for k = 1..6
wss <- map_dbl(1:6, function(k) {
  kmeans(clust_feats, centers = k, nstart = 10)$tot.withinss
})

tibble(k = 1:6, tot_withinss = wss)

# pick k = 3 (as proposed) and fit
kmeans_fit <- kmeans(clust_feats, centers = 3, nstart = 25)
kmeans_fit$centers

# Assign cluster labels back to train_data (first rows aligned)
train_clustered <- train_data %>%
  filter(row_number() <= nrow(clust_feats)) %>%
  mutate(cluster = factor(kmeans_fit$cluster))

# z-score outliers on returns
train_data <- train_data %>%
  mutate(z_return = (logret - mean(logret, na.rm = TRUE)) / sd(logret, na.rm = TRUE),
         outlier = abs(z_return) > 3)
table(train_data$outlier)
```

## 16. Model comparison table & backtest

```{r compare}
# Collate predictions for test set: all regression models
lm_pred <- predict(lm_fit, new_data = test_data) %>% pull(.pred)
rf_pred <- predict(rf_fit, new_data = test_data) %>% pull(.pred)
xgb_pred <- predict(xgb_fit, new_data = test_data) %>% pull(.pred)
dt_pred <- predict(dt_fit, new_data = test_data) %>% pull(.pred)

result_df <- test_data %>%
  select(Date, Close, Close_t1, return_1d) %>%
  mutate(
    pred_naive = lag(Close),
    pred_arima = test_arima$pred_close_arima,
    pred_lm = lm_pred,
    pred_rf = rf_pred,
    pred_xgb = xgb_pred,
    pred_dt = dt_pred
  ) %>%
  drop_na(pred_naive, pred_arima, pred_lm, pred_rf, pred_xgb, pred_dt)

# Regression metrics for all models
reg_metrics <- metric_set(rmse, mae, rsq)
cat("\nREGRESSION MODEL PERFORMANCE\n")
bind_rows(
  reg_metrics(result_df, truth = Close_t1, estimate = pred_naive) %>% mutate(model = "Naive"),
  reg_metrics(result_df, truth = Close_t1, estimate = pred_arima) %>% mutate(model = "ARIMA"),
  reg_metrics(result_df, truth = Close_t1, estimate = pred_lm) %>% mutate(model = "Linear Regression"),
  reg_metrics(result_df, truth = Close_t1, estimate = pred_rf) %>% mutate(model = "Random Forest"),
  reg_metrics(result_df, truth = Close_t1, estimate = pred_xgb) %>% mutate(model = "XGBoost"),
  reg_metrics(result_df, truth = Close_t1, estimate = pred_dt) %>% mutate(model = "Decision Tree")
) %>% print()

# Simple directional backtest (no transaction costs)
result_df <- result_df %>%
  mutate(
    sig_lm  = if_else(pred_lm > Close, 1, -1),
    sig_rf  = if_else(pred_rf > Close, 1, -1),
    sig_xgb = if_else(pred_xgb > Close, 1, -1),
    sig_dt  = if_else(pred_dt > Close, 1, -1),
    ret_lm  = sig_lm * return_1d,
    ret_rf  = sig_rf * return_1d,
    ret_xgb = sig_xgb * return_1d,
    ret_dt  = sig_dt * return_1d
  )

cat("\nBACKTEST RESULTS\n")
tibble(
  model = c("Linear Regression", "Random Forest", "XGBoost", "Decision Tree"),
  cumulative_return = c(
    prod(1 + na.omit(result_df$ret_lm)) - 1,
    prod(1 + na.omit(result_df$ret_rf)) - 1,
    prod(1 + na.omit(result_df$ret_xgb)) - 1,
    prod(1 + na.omit(result_df$ret_dt)) - 1
  ),
  directional_accuracy = c(
    mean((result_df$pred_lm > result_df$Close) == (result_df$Close_t1 > result_df$Close)),
    mean((result_df$pred_rf > result_df$Close) == (result_df$Close_t1 > result_df$Close)),
    mean((result_df$pred_xgb > result_df$Close) == (result_df$Close_t1 > result_df$Close)),
    mean((result_df$pred_dt > result_df$Close) == (result_df$Close_t1 > result_df$Close))
  )
) %>% print()

# Visualize model predictions vs actual (top 4 models)
result_plot_df <- result_df %>%
  select(Date, Close_t1, pred_lm, pred_rf, pred_xgb, pred_dt) %>%
  pivot_longer(cols = c(Close_t1, pred_lm, pred_rf, pred_xgb, pred_dt), names_to = "Type", values_to = "Price")

ggplot(result_plot_df, aes(x = Date, y = Price, color = Type)) +
  geom_line(size = 0.8, alpha = 0.7) +
  scale_color_manual(
    values = c("Close_t1" = "black", "pred_lm" = "blue", "pred_rf" = "red", "pred_xgb" = "purple", "pred_dt" = "orange"),
    labels = c("Actual", "Linear", "Random Forest", "XGBoost", "Decision Tree")
  ) +
  labs(
    title = "Model Predictions vs Actual Gold Prices",
    subtitle = "Test Set Performance - All Models",
    x = "Date",
    y = "Price (USD)",
    color = "Model"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Model performance comparison
perf_comparison <- tibble(
  Model = c("Naive", "ARIMA", "Linear Reg", "Random Forest", "XGBoost", "Decision Tree"),
  RMSE = c(
    rmse_vec(result_df$Close_t1, result_df$pred_naive),
    rmse_vec(result_df$Close_t1, result_df$pred_arima),
    rmse_vec(result_df$Close_t1, result_df$pred_lm),
    rmse_vec(result_df$Close_t1, result_df$pred_rf),
    rmse_vec(result_df$Close_t1, result_df$pred_xgb),
    rmse_vec(result_df$Close_t1, result_df$pred_dt)
  ),
  MAE = c(
    mae_vec(result_df$Close_t1, result_df$pred_naive),
    mae_vec(result_df$Close_t1, result_df$pred_arima),
    mae_vec(result_df$Close_t1, result_df$pred_lm),
    mae_vec(result_df$Close_t1, result_df$pred_rf),
    mae_vec(result_df$Close_t1, result_df$pred_xgb),
    mae_vec(result_df$Close_t1, result_df$pred_dt)
  )
)

perf_long <- perf_comparison %>%
  pivot_longer(cols = c(RMSE, MAE), names_to = "Metric", values_to = "Value")

ggplot(perf_long, aes(x = reorder(Model, -Value), y = Value, fill = Model)) +
  geom_col() +
  facet_wrap(~Metric, scales = "free_y") +
  coord_flip() +
  labs(
    title = "Model Performance Comparison (Lower is Better)",
    x = "Model",
    y = "Error Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Cumulative returns plot
result_df_cum <- result_df %>%
  mutate(
    cum_ret_lm = cumprod(1 + ret_lm),
    cum_ret_rf = cumprod(1 + ret_rf),
    cum_ret_xgb = cumprod(1 + ret_xgb),
    cum_ret_dt = cumprod(1 + ret_dt),
    cum_ret_bh = cumprod(1 + return_1d)  # buy and hold
  ) %>%
  select(Date, cum_ret_lm, cum_ret_rf, cum_ret_xgb, cum_ret_dt, cum_ret_bh) %>%
  pivot_longer(cols = c(cum_ret_lm, cum_ret_rf, cum_ret_xgb, cum_ret_dt, cum_ret_bh), names_to = "Strategy", values_to = "Cumulative_Return")

ggplot(result_df_cum, aes(x = Date, y = Cumulative_Return, color = Strategy)) +
  geom_line(size = 1) +
  scale_color_manual(
    values = c("cum_ret_lm" = "blue", "cum_ret_rf" = "red", "cum_ret_xgb" = "purple", "cum_ret_dt" = "orange", "cum_ret_bh" = "green"),
    labels = c("Linear", "Random Forest", "XGBoost", "Decision Tree", "Buy & Hold")
  ) +
  labs(
    title = "Cumulative Returns: Model Strategies vs Buy & Hold",
    x = "Date",
    y = "Cumulative Return (starting at 1)",
    color = "Strategy"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The Buy & Hold strategy (Green line) significantly outperformed all trading strategies. The Linear Regression strategy lost 48.9% and Random Forest lost 69.5%, indicating that while Linear Regression predicts price accurately, it does not necessarily predict *direction* (up/down) accurately enough for profitable trading.

## 17. Feature importance (RF permutation)

```{r importance}
# Extract the ranger model and get variable importance
rf_parsnip <- extract_fit_parsnip(rf_fit)$fit

# Get variable importance using ranger's built-in importance
importance_df <- tibble(
  Variable = names(rf_parsnip$variable.importance),
  Importance = rf_parsnip$variable.importance
) %>%
  arrange(desc(Importance))

cat("\nRANDOM FOREST VARIABLE IMPORTANCE \n")
print(importance_df)

# Plot importance
ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Random Forest Feature Importance",
    subtitle = "Permutation-based importance scores",
    x = "Features",
    y = "Importance"
  ) +
  theme_minimal()
```

## 18. Residual diagnostics (RF model)

```{r residuals}
train_preds <- predict(rf_fit, new_data = train_data) %>% bind_cols(train_data %>% select(Date, Close_t1))
train_resid <- train_preds %>% mutate(resid = Close_t1 - .pred)

ggplot(train_resid, aes(Date, resid)) +
  geom_line(color = "red", alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Random Forest Residuals Over Time",
    x = "Date",
    y = "Residual"
  ) +
  theme_minimal()

acf(na.omit(train_resid$resid), main = "ACF of Random Forest Residuals")
```
